### 2012-09-29 doc/prelim/document/chapters/introduction/introduction.tex
finished intro draft

We discuss in this chapter our motivation for developing these Monte
Carlo solvers by providing multiphysics problems of interest to the
nuclear engineering community. Hardware-based motivations are also
provided with the impact of forthcoming computing architectures
considered. In addition, background on the current state of solver
techniques for multiphysics problems and a brief comparision to the
proposed methods is provided to further motivate this work.

@@

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational with the Department
of Energy specifically allocating funding for the 2013 fiscal year to
research resilient solver technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}.

@@

The Forward-Automated Newton-MCSA (FANM) method is proposed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differencing (FAD) methods
\citep{bartlett_automatic_2006}, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated by generating the Jacobian explicitly with the model
equations through FAD. Second, the Jacobian need not be explicitly
formed by the user but is instead automated through FAD; this
eleminates the complexity of hand-coding derivatives and has also been
demonstrated to be more efficient computationally than evaluating
differenced derivatives. Third, unlike GMRES, MCSA does not build a
subspace during iterations. Although the Jacobian must be explicitly
formed to use MCSA, for problems that take more than a few GMRES
iterations to converge the size of the Krylov subspace will grow
beyond that of the Jacobian. Finally, using MCSA for the linear solve
provides its benefits for preconditioning, potential resiliency, and
parallelism.

This preliminary report oultines the conventional methods used in
practice for solving linear and nonlinear problems to provide a
mathematical basis upon which to build new algorithms that aim to
solve some of the aforementioned issues. These new algorithms will be
outlined in full with links made to past work and their potential for
offering improvements to the computational physics community. Progress
to date will be reported including the development of a new physics
framework that will serve as a test bed for these new methods that
leverages a preliminary implementation of the stochastic method. A
research plan will then be provided that continues the development and
implementation of these new methods and verifies them with benchmark
solutions, culminating in the analysis of a multiphysics simulation of
a pressurized water reactor assembly.

>>>>

We discuss in this chapter motivation for advancing Monte Carlo
solvers by providing multiphysics problems of interest in nuclear
reactor analysis. Hardware-based motivations are also provided by
considering the impact of forthcoming computing architectures. In
addition, background on the current state solver techniques for
multiphysics problems and a brief comparision to the proposed methods
is provided to further motivate this work.

@@

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational with the Department
of Energy's Advanced Scientific Computing Reasarch office specifically
allocating funding for the 2013 fiscal year to research resilient
solver technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues that a
set of robust, massively parallel stochastic solvers could provide a
solution for. As machines begin to operate at hundreds of petaflops
peak performance and beyond, restrictions on power avaiablity will
require incredibly high levels of concurrency. Furthermore, this drop
in power consumption will mean increased pressure on memory as memory
per node is expected to stagnate while cores per node is expected to
increase. As the number of cores increases, their clock speed is
expected to stagnate or even decrease to further reduce power
consumption and manufacturing costs.

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
overcome these events. With linear and nonlinear solvers based on
Monte Carlo techniques, such issues are alleviated by statistical
arguments. In the case of soft failures, isolated floating point
errors in Monte carlo simulation are absorbed within tally statistics
while completely losing hardware during a hard failure is manifested
as a high variance event where some portion of the Monte Carlo
histories are lost. These stochastic methods are a paradigm shift from
current determinstic solver techniques that will suffer greatly from
the non-deterministic behavior expected from these exascale machines.

In addition to resiliency concerns, the memory restrictions on future
hardware will hinder modern solvers that derive their robustness from
using large amounts of memory. Stochastic methods that are formulated
to use less memory than conventional methods will serve to alleviate
some of this pressure. In addition, new parallel strategies that may
be implemented with stochastic methods could offer a new avenue for
leveraging the expected levels of high concurrency in exascale
machines.

@@

The Forward-Automated Newton-MCSA (FANM) method is proposed as new
nonlinear solution method. The key features of FANM are: full Jacobian
generation using modern Forward Automated Differentiation (FAD)
methods \citep{bartlett_automatic_2006}, and MCSA as the inner linear
solver. This method has several attractive properties. First, the
first-order approximation to the Jacobian used in JFNK type methods is
eliminated by generating the Jacobian explicitly with the model
equations through FAD. Second, the Jacobian need not be explicitly
formed by the user but is instead automated through FAD; this
eleminates the complexity of hand-coding derivatives and has also been
demonstrated to be more efficient computationally than evaluating
differenced derivatives. Third, unlike GMRES, MCSA does not build a
subspace during iterations. Although the Jacobian must be explicitly
formed to use MCSA, for problems that take more than a few GMRES
iterations to converge the size of the Krylov subspace will grow
beyond that of the Jacobian. Finally, using MCSA for the linear solve
provides its benefits for preconditioning, potential resiliency, and
parallelism.

This preliminary report oultines in Chapter~\ref{ch:linear_problem}
the conventional methods used in practice for solving linear and
nonlinear problems to provide a mathematical basis upon which to build
new algorithms that aim to solve some of the aforementioned
issues. Parallel schemes for conventional methods are also provided
for background and understanding of how current methods may or may not
map to future hardware. In addition some components of their parallel
implementations may be applied to stochastic methods development. In
Chapter~\ref{ch:stochastic_methods}, Monte Carlo algorithms for
solving linear systems and new parallel strategies will be outlined in
full with links made to past work and their potential for offering
improvements to the multiphysics analysis community. From these
stochastic methods, a new nonlinear method is proposed in
Chapter~\ref{ch:nonlinear_problem} and compared to conventional
methods for solving nonlinear problems. The research proposal for this
work is presented in Chapter~\ref{ch:research_proposal} with progress
to date reported including the development of a new finite
element-based physics framework that leverages a preliminary general
implementation of the stochastic method that will serve as a test bed
for these new methods. An outline of a set of numerical experiments
will then be provided that continues the development and
implementation of these new methods and verifies them with benchmark
solutions, culminating in the analysis of a multiphysics simulation of
a pressurized water reactor subsystem using a production code system.
